A point-level trend model for the North American Breeding Bird Survey that corrects for detectability

### 2024-08-30 article/article.tex
thesis format

\documentclass[12pt]{article}

\usepackage{setspace}
\doublespacing

\usepackage{natbib}
\bibliographystyle{apalike}

\usepackage[margin=1.0in]{geometry}

\usepackage{amsmath}

\usepackage{graphicx}

\usepackage{array}
\usepackage{longtable}

\usepackage{lineno}
\linenumbers

\renewcommand{\topfraction}{.85}
\renewcommand{\bottomfraction}{.7}
\renewcommand{\textfraction}{.15}
\renewcommand{\floatpagefraction}{.66}
\renewcommand{\dbltopfraction}{.66}
\renewcommand{\dblfloatpagefraction}{.66}

%opening
\title{A point-level trend model for the North American Breeding Bird Survey that corrects for detectability}
\author{
	Edwards, Brandon P.M.\\
	\and
	Johnston, Alison\\
	\and
	Miller, David L.\\
	\and
	Bennett, Joseph R.\\
	\and
	Smith, Adam C.\\
}

\begin{document}

	\maketitle

	%\begin{abstract}

%	\end{abstract}

@@

\begin{figure}[h]
	\includegraphics{abundance-detectability_scaled.png}
	\caption{Toy example of how changing forest coverage can affect both true abundance and observed abundance of a particular forest species. In Year 1, with high forest coverage, there may be ten individuals at a patch but only four may be detectable due to the high forest coverage. Ten years later, because of forest loss, there may only be four individuals, and three of them can be detected. This may appear as only a slight population decrease, when in reality the population decreased from ten to four at this site.}
	\label{fig:example}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../output/plots/OVEN-map.png}
	\caption{Canada-wide trends for Ovenbird as derived from the status-quo route model in Equation \ref{bbs} and \citet{smith_spatially_2023} (A), the point-level model as in Equation \ref{bbs-point} (B), the point-level model that includes detectability as in Equation \ref{bbs-detect} (C), and the point-level model that includes detectability and variance propagation as in Equation \ref{bbs_varprop} (D).}
	\label{fig:oven-map}
\end{figure}

\begin{figure}[h]
	\includegraphics{../output/plots/OVEN-trajectory.png}
	\caption{Canada-wide trajectories for Ovenbird as derived from the status-quo route model in Equation \ref{bbs} and \citet{smith_spatially_2023} (A), the point-level model as in Equation \ref{bbs-point} (B), the point-level model that includes detectability as in Equation \ref{bbs-detect} (C), and the point-level model that includes detectability and variance propagation as in Equation \ref{bbs_varprop} (D). Be aware of the differences in the y-axes for each of these models: the ROUTE y-axis is the annual index of abundance per route, the POINT y-axis is the annual index of abundance per point, and the DETECT and VARPROP y-axis is the expected annual index of abundance per point given perfect detectability.}
	\label{fig:oven-trajectories}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../output/plots/route-vs-point.png}
	\caption{Index of abundance (A) and trend (B) derived from POINT model versus ROUTE model. Each point in A represents an index of abundance at the midyear in each stratum, whereas each point in B represents a trend estimate for a stratum. Regression lines (black) and posterior draws (grey) indicate modelled relationship.}
	\label{fig:route-vs-point}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../output/plots/precision_comparison.png}
	\caption{Boxplot of 90\% credible interval widths for trend estimates from each model.}
	\label{fig:precision}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../output/plots/varprop-diagnostics.png}
	\caption{Diagnostics of variance propagation as suggested by \citet{bravington_variance_2021}. A) Index of abundance as derived from the DETECT model versus index of abundance as derived from the VARPROP model, from the midyear from each stratum (points). Regression line (black) and posterior draws (grey) indicate modelled relationship. B) Variance propagation parameter posterior distributions for availability. Because we are using the ``best" NA-POPS removal model for Ovenbird, we only have parameter distributions for the Intercept, Ordinal Day, and Ordinal Day squared terms. C) Variance propagation parameter posterior distributions for perceptibility. Similar to B), because we are using the ``best" NA-POPS distance model for Ovenbird, we include distributions for the Intercept, Forest, Road, and interaction terms.}
	\label{fig:varprop-diagnostics}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../output/plots/trend-vs-forest.png}
	\caption{Change in forest coverage versus difference in trend between the VARPROP and POINT models. Change in forest coverage was calculated as the difference between forest coverage in 2022 and 2010. Trend difference was calculated as the VARPROP trend minus the POINT trend. Regression line (black) and posterior draws (grey) indicate the modelled relationship.}
	\label{fig:trend-vs-forest}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../output/plots/route-level-change.png}
	\caption{A) Map of forest coverage change from 2010 - 2022 at all Breeding Bird Survey points in Canada for Ovenbird. Inset is a particular route of interest in Ontario that has experienced significant forest coverage loss from 2010 - 2022. For this route, we compared the population trajectories of Ovenbird using the POINT model (B) to the population trajectories of Ovenbird using the VARPROP model (C) to assess differences in trend.}
	\label{fig:route-change}
\end{figure}

\begin{figure}[h]
	\includegraphics{../output/plots/OVENpoint-vs-detectability-map.png}
	\caption{Difference in trend between the VARPROP and POINT model (i.e., VARPROP - POINT; A) and the DETECT and POINT model (i.e., DETECT - POINT; B).}
	\label{fig:route-change}
\end{figure}

@@

	\bibliography{refs}

\end{document}

>>>>

\chapter{A point-level trend model for the North American Breeding Bird Survey that corrects for detectability}

\section{Abstract}
\par TO DO

@@

\begin{figure}[h]
	\includegraphics{../ch5-bbs-detectability/article/abundance-detectability_scaled.png}
	\caption{Toy example of how changing forest coverage can affect both true abundance and observed abundance of a particular forest species. In Year 1, with high forest coverage, there may be ten individuals at a patch but only four may be detectable due to the high forest coverage. Ten years later, because of forest loss, there may only be four individuals, and three of them can be detected. This may appear as only a slight population decrease, when in reality the population decreased from ten to four at this site.}
	\label{fig:example}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../ch5-bbs-detectability/output/plots/OVEN-map.png}
	\caption{Canada-wide trends for Ovenbird as derived from the status-quo route model in Equation \ref{bbs} and \citet{smith_spatially_2023} (A), the point-level model as in Equation \ref{bbs-point} (B), the point-level model that includes detectability as in Equation \ref{bbs-detect} (C), and the point-level model that includes detectability and variance propagation as in Equation \ref{bbs_varprop} (D).}
	\label{fig:oven-map}
\end{figure}

\begin{figure}[h]
	\includegraphics{../ch5-bbs-detectability/output/plots/OVEN-trajectory.png}
	\caption{Canada-wide trajectories for Ovenbird as derived from the status-quo route model in Equation \ref{bbs} and \citet{smith_spatially_2023} (A), the point-level model as in Equation \ref{bbs-point} (B), the point-level model that includes detectability as in Equation \ref{bbs-detect} (C), and the point-level model that includes detectability and variance propagation as in Equation \ref{bbs_varprop} (D). Be aware of the differences in the y-axes for each of these models: the ROUTE y-axis is the annual index of abundance per route, the POINT y-axis is the annual index of abundance per point, and the DETECT and VARPROP y-axis is the expected annual index of abundance per point given perfect detectability.}
	\label{fig:oven-trajectories}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../ch5-bbs-detectability/output/plots/route-vs-point.png}
	\caption{Index of abundance (A) and trend (B) derived from POINT model versus ROUTE model. Each point in A represents an index of abundance at the midyear in each stratum, whereas each point in B represents a trend estimate for a stratum. Regression lines (black) and posterior draws (grey) indicate modelled relationship.}
	\label{fig:route-vs-point}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../ch5-bbs-detectability/output/plots/precision_comparison.png}
	\caption{Boxplot of 90\% credible interval widths for trend estimates from each model.}
	\label{fig:precision}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../ch5-bbs-detectability/output/plots/varprop-diagnostics.png}
	\caption{Diagnostics of variance propagation as suggested by \citet{bravington_variance_2021}. A) Index of abundance as derived from the DETECT model versus index of abundance as derived from the VARPROP model, from the midyear from each stratum (points). Regression line (black) and posterior draws (grey) indicate modelled relationship. B) Variance propagation parameter posterior distributions for availability. Because we are using the ``best" NA-POPS removal model for Ovenbird, we only have parameter distributions for the Intercept, Ordinal Day, and Ordinal Day squared terms. C) Variance propagation parameter posterior distributions for perceptibility. Similar to B), because we are using the ``best" NA-POPS distance model for Ovenbird, we include distributions for the Intercept, Forest, Road, and interaction terms.}
	\label{fig:varprop-diagnostics}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../ch5-bbs-detectability/output/plots/trend-vs-forest.png}
	\caption{Change in forest coverage versus difference in trend between the VARPROP and POINT models. Change in forest coverage was calculated as the difference between forest coverage in 2022 and 2010. Trend difference was calculated as the VARPROP trend minus the POINT trend. Regression line (black) and posterior draws (grey) indicate the modelled relationship.}
	\label{fig:trend-vs-forest}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../ch5-bbs-detectability/output/plots/route-level-change.png}
	\caption{A) Map of forest coverage change from 2010 - 2022 at all Breeding Bird Survey points in Canada for Ovenbird. Inset is a particular route of interest in Ontario that has experienced significant forest coverage loss from 2010 - 2022. For this route, we compared the population trajectories of Ovenbird using the POINT model (B) to the population trajectories of Ovenbird using the VARPROP model (C) to assess differences in trend.}
	\label{fig:route-change}
\end{figure}

\begin{figure}[h]
	\includegraphics{../ch5-bbs-detectability/output/plots/OVENpoint-vs-detectability-map.png}
	\caption{Difference in trend between the VARPROP and POINT model (i.e., VARPROP - POINT; A) and the DETECT and POINT model (i.e., DETECT - POINT; B).}
	\label{fig:route-change}
\end{figure}

@@
